Escaping RGBland: Selecting Colors for Statistical Graphics

### 2009-02-11 pkg/paper/Palettes/CSDA.Rnw
Re-arranged graphics for CSDA (saving 5 x EUR 473!!).
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@53 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[shortnames,noheadings]{Z}

\usepackage{amsfonts}
\usepackage{rotating}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}

@@

  png(file = "bkde.png", height = 800, width = 830)
  par(mfrow = c(2, 2), mar = c(4.5, 3.5, 1, 1))
  plot_bkde2D(dens, col = hsv(h = seq(from = 2/3, to = 0, length = 33), v = 0.9, s = 0.9))
  plot_bkde2D(dens, col = rev(heat_hcl(33, h = c(0, 90), c = 0,         l = c(30, 90), power = c(1/5, 1.3))))
  plot_bkde2D(dens, col = rev(heat.colors(33)))
  plot_bkde2D(dens, col = rev(heat_hcl(33, h = c(0, 90), c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
  dev.off()
}
@
\includegraphics{bkde}
\setkeys{Gin}{width=.47\textwidth}
<<bkde-pal>>=
if(generate_png) {
  png(file = "bkde-pal.png", height = 160, width = 400)
  par(mfrow = c(4, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.7, 0, 1.9))
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
    hsv(h = seq(from = 2/3, to = 0, length = 99), v = 0.9, s = 0.9))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
    rev(heat.colors(99)))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
    rev(heat_hcl(99, h = c(0, 90), c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
    rev(heat_hcl(99, h = c(0, 90), c = 0,         l = c(30, 90), power = c(1/5, 1.3))))
  box()
  axis(1, cex.axis = 1.8)
  dev.off()
}
@
\includegraphics{bkde-pal}
\caption{\label{fig:bkde} Bivariate density estimation of duration (x-axis) and
waiting time (y-axis) for Old Faithful geyser eruptions. The palettes employed
are (counterclockwise) an HSV-based rainbow, HSV-based heat colors, HCL-based heat colors
and grayscales.}
\end{center}
\end{figure}

@@

  png(file = "nigeria.png", height = 1050, width = 840)
  par(mfrow = c(3, 2), mar = c(1.22, 0.82, 1.02, 0.82))
  plot(m, z = fit$pmode, fill =           rev(rainbow(99, start = 0, end = 1/3)),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, fill = dichromat(rev(rainbow(99, start = 0, end = 1/3)), type = "protan"),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill =           rev(diverge_hcl(99, h = c(0, 120),   c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(0, 120),   c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill =           rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  dev.off()
}
@
\includegraphics{nigeria}
\setkeys{Gin}{width=.47\textwidth}
<<nigeria-pal>>=
if(generate_png) {
  png(file = "nigeria-pal.png", height = 120, width = 396)
  par(mfrow = c(3, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 0.2, 0, 0.2))
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = 
    rev(rainbow(99, start = 0, end = 1/3)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = 
    rev(diverge_hcl(99, h = c(0, 120),   c = 60, l = c(45, 90), power = 1.2)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = 
    rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  box()
  axis(1, cex.axis = 1.8)
  dev.off()
}
@
\includegraphics{nigeria-pal}
\caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.
The color palettes employed are an HSV-based rainbow and two HCL-based diverging palettes.
In the right panels red-green contrasts are collapsed to emulate protanopic vision.}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<wheel, fig=TRUE, height=3.6, width=10>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
par(mfrow = c(1, 2), mar = c(1, 1, 1, 1), oma = c(0, 0, 0, 0))
par(lwd = 3)
pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6, init.angle = -30)
par(lwd = 0.5, new = TRUE)
pie(rep(1, 90), radius = 1, col = rainbow(90), border = NA, labels = NA, init.angle = -2)
par(lwd = 3, new = TRUE)
pie(1, radius = 0.55, labels = NA)
par(lwd = 0.5, new = TRUE)
pie(1, radius = 0.55, border = NA, labels = NA)
par(lwd = 3)
pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6, init.angle = -30)
par(lwd=0.5, new=TRUE)
pie(rep(1, 90), radius = 1, col = rainbow_hcl(90, c = 50, l = 70), border = NA, labels = NA, init.angle = -2)
par(lwd = 3, new = TRUE)
pie(1, radius = 0.55, labels = NA)
par(lwd = 0.5, new = TRUE)
pie(1, radius = 0.55, border = NA, labels = NA)
@
\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-q-construct>>=
if(generate_png) {
  png(file = "pal-q-construct.png", height = 480, width = 480)
  hcl.wheel(luminance = 70, do.label = FALSE, inner.circle = TRUE)
  rect(0.45, -0.05, 0.55, 0.05, col = hcl2hex(0, 50, 70))
  rect(-cos(pi/3)/2-0.05, sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, sin(pi/3)/2+0.05, col = hcl2hex(120, 50, 70))
  rect(-cos(pi/3)/2-0.05, -sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, -sin(pi/3)/2+0.05, col = hcl2hex(240, 50, 70))
  dev.off()
}
@
\includegraphics{pal-q-construct}
\vspace*{-2cm}
\caption{\label{fig:pal-q-construct} Constructing qualitative palettes. In the hue/chroma
plane for $L = 70$, the dashed circle correponds to a radius $C = 50$ with chosen angles
$H = 0, 120, 240$.}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-q-examples, fig=TRUE, height=4, width=6>>=
par(mfrow = c(2, 2), mar = c(0, 0, 3, 2))
pal(rainbow_hcl(4, start =  30, end = 300, c = 50, l = 70), main = "dynamic [30, 300]")
pal(rainbow_hcl(4, start =  60, end = 240, c = 50, l = 70), main = "harmonic [60, 240]")
pal(rainbow_hcl(4, start = 270, end = 150, c = 50, l = 70), main = "cold [270, 150]")
pal(rainbow_hcl(4, start =  90, end = -30, c = 50, l = 70), main = "warm [90, -30]")
@
\caption{\label{fig:pal-q-examples} Examples for qualitative palettes. Hue is varied
in different intervals for given $C = 50$ and $L = 70$.}
\end{center}
\end{figure}

@@

Figure~\ref{fig:pal-q-construct} depicts how the HCL-based color wheel is constructed.
It shows the hue/chroma plane of HCL space given a luminance of $L = 70$. Not all
combinations of hue and chroma are admissible, however for a chroma of $C = 50$ a full
color wheel can be obtained.
For choosing the hues in a certain palette, various strategies are conceivable.
A simple and intuitive one is to use colors as metaphors for categories (e.g., for political parties),
%% (e.g., for political parties as in Figures~\ref{fig:seats} and \ref{fig:votes}),
another approach would be to use segments from
the color wheel corresponding to nearby or distant colors. The latter is shown
in Figure~\ref{fig:pal-q-examples} which depicts
examples for generating qualitative sets of colors $(H, 50, 70)$.
In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
set with $H = 60, 120, 180, 240$. Cold colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and warm colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.

@@

Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
  \[ (H, 0 + i' \cdot C_{\max}, L_{\max} - i' \cdot (L_{\max} - L_{\min})). \]
This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
hue $H = 260$ with $L_{\max} = 90$
and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
The first two combinations are also shown in the first two rows of Figure~\ref{fig:pal-s-examples}.
The right panel of Figure~\ref{fig:pal-s-construct} shows that the exact same strategy
is not possible  for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
the same way, there is a stronger trade-off between using dark colors (with low luminance)
and colorful colors (with high chroma). Hence, the second path from light gray to 
full green ends at a much lighter color with $L_{\min} = 75$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-construct>>=
if(generate_png) {
  png(file = "pal-s-construct.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  hue.slice(260, grid.n = 301)
  lines(c(0, 100), c(90, 50), lty = 2)
  lines(c(0, 80), c(90, 30), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
  rect(47.5, 67.5, 52.5, 72.5, col = hcl2hex(260, 50, 70))
  rect(95, 47.5, 100, 52.5, col = hcl2hex(260, 100, 50))
  rect(37.5, 57.5, 42.5, 62.5, col = hcl2hex(260, 40, 60))
  rect(77.5, 27.5, 82.5, 32.5, col = hcl2hex(260, 80, 30))
  rect(0, 57.5, 5, 62.5, col = hcl2hex(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl2hex(260, 0, 30))
  hue.slice(120, grid.n = 301)
  lines(c(0, 100), c(90, 75), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
  rect(47.5, 80, 52.5, 85, col = hcl2hex(120, 50, 82.5))
  rect(95, 72.5, 100, 77.5, col = hcl2hex(120, 100, 75))
  rect(0, 57.5, 5, 62.5, col = hcl2hex(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl2hex(260, 0, 30))
  dev.off()
}
@
\includegraphics{pal-s-construct}
\caption{\label{fig:pal-s-construct} Constructing sequential palettes. The chroma/luminance
plane is shown for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
either only luminance or both luminance and chroma.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0.25, 0, 0.25, 0))
pal(sequential_hcl(12, h = 260,        c = 0,         l = c(30, 90), power = 2.2))
pal(sequential_hcl(12, h = 260,        c = c(80, 0),  l = c(30, 90), power = 2.2))
pal(      heat_hcl(12, h = c(0, 90),   c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
pal(rev(  heat_hcl(12, h = c(0, -100), c = c(40, 80), l = c(75, 40), power = 1)))
pal(   terrain_hcl(12, h = c(130, 0),  c = c(65, 0),  l = c(45, 90), power = c(1/2, 1.5)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}

To increase the contrast between the colors in the palette even further, the ideas from
the previous sequential palettes can also be combined with qualitative palettes by
simultaneously varying the hue as well:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i' \cdot (C_{\max} - C_{\min}),
      L_{\max} - i'' \cdot (L_{\max} - L_{\min})). \]
One application is an HCL-based version of ``heat colors'' that increase from a light yellow
(e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
visible, the chroma needs to increase rather quickly for low values of $i$ and then only
slowly for higher values of $i$. This can be achieved by choosing an appropriate transformation
$i'$ for chroma and a different transformation $i''$ for the luminance.
Such a strategy is adopted for the palettes shown in the lower three rows in
Figure~\ref{fig:pal-s-examples} using different pairs of hues as well as different
chroma and luminance contrasts.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<seats, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
names(parties_hcl) <- names(parties_hsv) <- names(seats)
names(seats)[4] <- "Grüne"
Encoding(names(seats)) <- "latin1"
pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

>>>>

\documentclass[shortnames,noheadings]{Z}
\usepackage{amsfonts,thumbpdf}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}

@@

  png(file = "bkde.png", height = 1560, width = 1640, res = 144)
  par(mfrow = c(2, 2), mar = c(4.5, 3.5, 1, 1))
  plot_bkde2D(dens, col = hsv(h = seq(from = 2/3, to = 0, length = 33), v = 0.9, s = 0.9))
  plot_bkde2D(dens, col = rev(heat_hcl(33, h = c(0, 90), c = 0,         l = c(30, 90), power = c(1/5, 1.3))))
  plot_bkde2D(dens, col = rev(heat.colors(33)))
  plot_bkde2D(dens, col = rev(heat_hcl(33, h = c(0, 90), c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
  dev.off()
}
@
\includegraphics{bkde}
\setkeys{Gin}{width=.47\textwidth}
<<bkde-pal>>=
if(generate_png) {
  png(file = "bkde-pal.png", height = 320, width = 800, res = 144)
  par(mfrow = c(4, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.9, 0, 1.9))
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
    hsv(h = seq(from = 2/3, to = 0, length = 99), v = 0.9, s = 0.9))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
    rev(heat.colors(99)))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
    rev(heat_hcl(99, h = c(0, 90), c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
    rev(heat_hcl(99, h = c(0, 90), c = 0,         l = c(30, 90), power = c(1/5, 1.3))))
  box()
  axis(1, cex.axis = 1.8)
  dev.off()
}
@
\includegraphics{bkde-pal}
\caption{\label{fig:bkde} Bivariate density estimation of duration (x-axis) and
waiting time (y-axis) for Old Faithful geyser eruptions. The palettes employed
are (counterclockwise from top left) an HSV-based rainbow, HSV-based heat colors, HCL-based heat colors
and grayscales.}
\end{center}
\end{figure}

@@

  png(file = "nigeria.png", height = 2100, width = 1680, res = 144)
  par(mfrow = c(3, 2), mar = c(1.22, 0.82, 1.02, 0.82))
  plot(m, z = fit$pmode, fill =           rev(rainbow(99, start = 0, end = 1/3)),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, fill = dichromat(rev(rainbow(99, start = 0, end = 1/3)), type = "protan"),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill =           rev(diverge_hcl(99, h = c(0, 120),   c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(0, 120),   c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill =           rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  dev.off()
}
@
\includegraphics{nigeria}
\setkeys{Gin}{width=.47\textwidth}
<<nigeria-pal>>=
if(generate_png) {
  png(file = "nigeria-pal.png", height = 240, width = 800, res = 144)
  par(mfrow = c(3, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.9, 0, 1.9))
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = 
    rev(rainbow(99, start = 0, end = 1/3)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = 
    rev(diverge_hcl(99, h = c(0, 120),   c = 60, l = c(45, 90), power = 1.2)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = 
    rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  box()
  axis(1, cex.axis = 1.8)
  dev.off()
}
@
\includegraphics{nigeria-pal}
\caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.
The color palettes employed are (from top to bottom) an HSV-based rainbow and two
HCL-based diverging palettes. In the right panels red-green contrasts are collapsed to
emulate protanopic vision.}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<wheel>>=
png(file = "wheel.png", height = 500, width = 1400, res = 144)
par(mar = rep(0.8, 4), mfrow = c(1, 2))
par(mfrow = c(1, 2), mar = c(1.1, 1, 1.1, 1), oma = c(0, 0, 0, 0))
par(lwd = 3)
pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6, init.angle = -30)
par(lwd = 0.5, new = TRUE)
pie(rep(1, 180), radius = 1, col = rainbow(180), border = NA, labels = NA, init.angle = -1)
par(lwd = 3, new = TRUE)
pie(1, radius = 0.55, labels = NA)
par(lwd = 0.5, new = TRUE)
pie(1, radius = 0.55, border = NA, labels = NA)
par(lwd = 3)
pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6, init.angle = -30)
par(lwd=0.5, new=TRUE)
pie(rep(1, 180), radius = 1, col = rainbow_hcl(180, c = 50, l = 70), border = NA, labels = NA, init.angle = -1)
par(lwd = 3, new = TRUE)
pie(1, radius = 0.55, labels = NA)
par(lwd = 0.5, new = TRUE)
pie(1, radius = 0.55, border = NA, labels = NA)
dev.off()
@
\includegraphics{wheel}
\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.55\textwidth}
\begin{figure}[t!]
\begin{center}
<<pal-q-construct>>=
if(generate_png) {
  png(file = "pal-q-construct.png", height = 960, width = 960, res = 144)
  par(mar = c(0, 0, 0, 0))
  hcl.wheel(luminance = 70, do.label = FALSE, inner.circle = TRUE)
  rect(0.45, -0.05, 0.55, 0.05, col = hcl2hex(0, 50, 70))
  rect(-cos(pi/3)/2-0.05, sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, sin(pi/3)/2+0.05, col = hcl2hex(120, 50, 70))
  rect(-cos(pi/3)/2-0.05, -sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, -sin(pi/3)/2+0.05, col = hcl2hex(240, 50, 70))
  dev.off()
}
@
\includegraphics{pal-q-construct}

\vspace*{1cm}

\setkeys{Gin}{width=0.75\textwidth}
<<pal-q-examples, fig=TRUE, height=3, width=6>>=
par(mfrow = c(2, 2), mar = c(1, 0, 3, 2))
pal(rainbow_hcl(4, start =  30, end = 300, c = 50, l = 70), main = "dynamic [30, 300]")
pal(rainbow_hcl(4, start =  60, end = 240, c = 50, l = 70), main = "harmonic [60, 240]")
pal(rainbow_hcl(4, start = 270, end = 150, c = 50, l = 70), main = "cold [270, 150]")
pal(rainbow_hcl(4, start =  90, end = -30, c = 50, l = 70), main = "warm [90, -30]")
@
\caption{\label{fig:pal-q} Qualitative palettes: construction and examples.
Top: Construction in the hue/chroma
plane for $L = 70$, the dashed circle correponds to a radius $C = 50$ with chosen angles
$H = 0, 120, 240$. Bottom: Examples by varying hue in different intervals for
given $C = 50$ and $L = 70$.}
\end{center}
\end{figure}

@@

Figure~\ref{fig:pal-q} (top panel) depicts how the HCL-based color wheel is constructed.
It shows the hue/chroma plane of HCL space given a luminance of $L = 70$. Not all
combinations of hue and chroma are admissible, however for a chroma of $C = 50$ a full
color wheel can be obtained.
For choosing the hues in a certain palette, various strategies are conceivable.
A simple and intuitive one is to use colors as metaphors for categories (e.g., for political parties),
another approach would be to use segments from
the color wheel corresponding to nearby or distant colors. The latter is shown
in Figure~\ref{fig:pal-q} (bottom panel) which depicts
examples for generating qualitative sets of colors $(H, 50, 70)$.
In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
set with $H = 60, 120, 180, 240$. Cold colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and warm colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.

\newpage

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<pal-s-construct>>=
if(generate_png) {
  png(file = "pal-s-construct.png", height = 840, width = 1680, res = 144)
  par(mfrow = c(1, 2))
  hue.slice(260, grid.n = 301)
  lines(c(0, 100), c(90, 50), lty = 2)
  lines(c(0, 80), c(90, 30), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
  rect(47.5, 67.5, 52.5, 72.5, col = hcl2hex(260, 50, 70))
  rect(95, 47.5, 100, 52.5, col = hcl2hex(260, 100, 50))
  rect(37.5, 57.5, 42.5, 62.5, col = hcl2hex(260, 40, 60))
  rect(77.5, 27.5, 82.5, 32.5, col = hcl2hex(260, 80, 30))
  rect(0, 57.5, 5, 62.5, col = hcl2hex(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl2hex(260, 0, 30))
  hue.slice(120, grid.n = 301)
  lines(c(0, 100), c(90, 75), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
  rect(47.5, 80, 52.5, 85, col = hcl2hex(120, 50, 82.5))
  rect(95, 72.5, 100, 77.5, col = hcl2hex(120, 100, 75))
  rect(0, 57.5, 5, 62.5, col = hcl2hex(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl2hex(260, 0, 30))
  dev.off()
}
@
\includegraphics{pal-s-construct}

\vspace*{1cm}

\setkeys{Gin}{width=.75\textwidth}
<<pal-s-examples, fig=TRUE, height=4, width=6>>=
par(mfrow = c(5, 1), mar = c(0.25, 0, 0.25, 0))
pal(sequential_hcl(12, h = 260,        c = 0,         l = c(30, 90), power = 2.2))
pal(sequential_hcl(12, h = 260,        c = c(80, 0),  l = c(30, 90), power = 2.2))
pal(      heat_hcl(12, h = c(0, 90),   c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
pal(rev(  heat_hcl(12, h = c(0, -100), c = c(40, 80), l = c(75, 40), power = 1)))
pal(   terrain_hcl(12, h = c(130, 0),  c = c(65, 0),  l = c(45, 90), power = c(1/2, 1.5)))
@

\caption{\label{fig:pal-s} Sequential palettes: construction and examples.
Top: Construction is shown in the chroma/luminance
plane for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
either only luminance or both luminance and chroma. Bottom: Examples by varying
only luminance (first row), chroma and luminance (second row), and hue,
chroma and luminance (remaining rows).}
\end{center}
\end{figure}

Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
  \[ (H, 0 + i' \cdot C_{\max}, L_{\max} - i' \cdot (L_{\max} - L_{\min})). \]
This strategy is depicted in the top left panel of Figure~\ref{fig:pal-s} for a blue
hue $H = 260$ with $L_{\max} = 90$
and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
The first two combinations are also shown in the first two rows of the bottom panel in Figure~\ref{fig:pal-s}.
The top right panel of Figure~\ref{fig:pal-s} shows that the exact same strategy
is not possible  for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
the same way, there is a stronger trade-off between using dark colors (with low luminance)
and colorful colors (with high chroma). Hence, the second path from light gray to 
full green ends at a much lighter color with $L_{\min} = 75$.

To increase the contrast between the colors in the palette even further, the ideas from
the previous sequential palettes can also be combined with qualitative palettes by
simultaneously varying the hue as well:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i' \cdot (C_{\max} - C_{\min}),
      L_{\max} - i'' \cdot (L_{\max} - L_{\min})). \]
One application is an HCL-based version of ``heat colors'' that increase from a light yellow
(e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
visible, the chroma needs to increase rather quickly for low values of $i$ and then only
slowly for higher values of $i$. This can be achieved by choosing an appropriate transformation
$i'$ for chroma and a different transformation $i''$ for the luminance.
Such a strategy is adopted for the palettes shown in the last three rows in
Figure~\ref{fig:pal-s} using different pairs of hues as well as different
chroma and luminance contrasts.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<seats, fig=TRUE, height=4, width=11>>=
par(mar = c(0.8, 1.6, 0, 0), mfrow = c(1, 2))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
names(parties_hcl) <- names(parties_hsv) <- names(seats)
names(seats)[4] <- "Grüne"
Encoding(names(seats)) <- "latin1"
pie(seats, clockwise = TRUE, col = parties_hsv, radius = 0.85)
pie(seats, clockwise = TRUE, col = parties_hcl, radius = 0.85)
@

<<votes, fig=TRUE, height=5.5, width=13>>=
data("Bundestag2005", package = "vcd")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
parties_hcl <- parties_hcl[colnames(votes)]
parties_hsv <- parties_hsv[colnames(votes)]
dimnames(votes)[[1]][c(9, 16)] <- c("Baden-Württemberg", "Thüringen")
dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
Encoding(dimnames(votes)[[1]]) <- "latin1"
